Generalised gravitational burst generation with Generative Adversarial Networks

### 2020-09-03 Paper_draft/main.tex
Added comments to section 1

\begin{abstract}
The next generation of Gravitational wave detectors will accelerate the number
of gravitational wave detections such that we can gain new in site into the
physics behind the sources causing the phenomena. Numerical simulations and
matched filtering are the standard for detecting gravitational waves for known
sources such as binary-black hole mergers. There are other sources of gravitational waves
that remain elusive to standard modelling techniques and are expected to be
detectable. Here we construct a unmodelled burst generation scheme using
Generative adversarial networks - a powerful class of machine learning.
\end{abstract}

@@

\acrodef{GW}[GW]{Gravitational wave}
\acrodef{GAN}[GAN]{generative adversarial network}
\acrodef{CGAN}[CGAN]{conditional generative adversarial network}
\acrodef{ACGAN}[ACGAN]{auxilliary conditional generative adversarial network}
\acrodef{DCGAN}[DCGAN]{deep convolutional generative adversarial network}
\acrodef{CNN}[CNN]{convolutional neural networks}
\acrodef{BBH}[BBH]{binary black hole}

@@

% introduce \ac{GW} astrophsyics
%
Gravitational wave astronomy is now an established field, starting with the first
detection of a binary black hole merger \cite{Abbott2016} in September 2015.
Following this, the first and second observations runs (O1 and O2) of Advanced
LIGO and Advanced Virgo \cite{Prospects-dets, AdvLIGO, AdvLIGO2, AdvVIRGO} reported several more Compact Binary Coalescence (CBC)
mergers \cite{Abbott2016a, Abbott2017, Abbott2017a, Abbott2017b}. On August
2017 a binary neutron star merger was observed alongside its electromagnetic
counterpart for the first time, giving rise to multimessenger gravitational
wave astronomy.

% introduce burst searches
%
With these successes and continued upgrades to the detectors, further detections of CBCs are expected to be commonplace in a few short years. Another group of GW signals that has thus far been undetectable is GW ``bursts".  GW bursts are transient signals of typically short duration ($<$ 1s) whose
waveforms are not accurately modelled or are complex to re-produce.
Astrophysical sources for such transients include: Core collapse supernova \cite{Fryer_2003}, Pulsar glitches \cite{Andersson_2001}, Neutron star
post-mergers \cite{Baiotti_2007} and other as-yet unexplained physical phenomena.

GW detections so far have used a process called matched filtering, \cite{Owen1998}, where a large template bank of possible GW waveforms are compared to the detector outputs. In order to increase the chances of detection these template banks must span a large multi dimensional parameter space and the templates must be accurate, requiring significant computational cost.  GW bursts are un-modelled; therefore there are no templates available and so these signals are undetectable through this scheme. Instead, detection involves distinguishing
the signal from detector noise. Burst searches look for excess power contained in the time-frequency domain and rely on the astrophysical burst waveform appearing in multiple detectors at similar times. This is only possible if the detector noise is
well characterised and the candidate signal can be differentiated from
systematic or environmental glitches.

% Discuss the family of burst waveforms currently used and why - not in detail, just
% an introduction 
%
\ac{GW} burst detection algorithms \cite{Klimenko_2008, Aso_2008} are tested and tuned using modelled waveforms that may
or may not have astrophysical significance but have easy to define parameters
and share characteristics of real bursts that are enough to simulate a \ac{GW} passing between detectors. Such waveforms include sine-Gaussians: a Gaussian modulated sine wave that is characterised by its central frequency and decay parameter. Bandlimited white-noise bursts: white noise that is contained within a certain frequency range and ring-downs which mimic the damped oscillations after a CBC merger.

\begin{comment}Such waveforms may have
long-duration, short bandwidth (ringdowns), long-duration, large bandwidth
(inspirals) and many algorithms make use of sine-Gaussians: a Gaussian
modulated sine wave that is characterised by it's central frequency and narrow
bandwidth.~\chris{not the time to try to describe the types of burst waveforms.
Also, be careful with satying things like ringdowns have narrow bandwidth. I
know we specify a single frequency but becuase of the short duration, the
signal is broad band. Look at the FFT} This makes it a great tool for
diagnosing LIGOs sensitivity to frequency.~\chris{strange unfinished sentence.} 
\end{comment}

% Introduce ML techniques in GWs
%

With the expectation that there will be many more \ac{GW} detections in the future, researchers are starting to realise the need for fast and efficient \ac{GW} analysis to compete with the rising number of detections. While still in its infancy, Machine Learning (ML) with GW has already shown great potential for overcoming data bottlenecks. \ac{GW} analysis in areas of detection \cite{Gabbard2017,Gebhard_2019,Krastev_2020}, glitch classification \cite{Bahaadini, George_2018,Razzano_2018} and
parameter estimation \cite{gabbard2019bayesian,shen2019deterministic,green2020gravitationalwave}. Applying these algorithms show similar results to matched filtering techniques and have the benefit of being orders of magnitude faster. \michael{We mention aspects related to detection (pattern matching) but also metion PE in the same sentence and I think this is a bit confusing.}

% What this paper does on GANs in 1 paragraph
%

In this work we aim to explore the use of machine learning to generate and interpret
unmodelled \ac{GW} burst waveforms. Using the generative machine learning model: Generative Adversarial Networks (GANs), we train on five classes of known burst morphologies in the time domain. Working with the assumption that GANs construct smooth n-dimensional vector spaces between its input and output, we can then explore the space between the five classes to construct new unmodelled waveforms. As all the computationally expensive processes occur during training, the learned model after
training is able to produce waveforms in fractions of a second and in fact produce waveforms that are impossible to produce with current techniques. These new varieties of waveforms can then be used to diagnose detection algorithms and gain new insight into sources of \ac{GW} bursts. 
% the point of the paper
%

This paper is organised as follows: \jordan{unstructured at the moment}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this work we present the potential of Generative Adversarial Networks for burst gravitational wave analysis. We have shown that GANs have the ability to generate 5 class varieties of modelled burst \ac{GW} signals that can be generated at whim. The latent and class spaces were explored through interpolation and suggest that the space provides smooth translations between classes and overall waveform shape. We then showed targeted waveform generation by mixing classes to produce new unmodlled waveform varieties that can be used to test current burst search pipelines. \jordan{couple of sentences about classifier}.

@@

The approach shown in this work shows promise in generating unmodelled burst waveforms from exotic sources. Having the ability to quickly generate new waveforms is essential to test current detection schemes and their susceptibilty to unmodelled sources. We belive that GANs have the ability to generate high fidelity waveforms at a fraction of the computational expense and do not rely on large prior parameter space. Having banks of these waveforms at hand can aid in our understand of the physics processes behind these non-standard gravitational wave emitters.

@@

\section*{References}
\bibliography{iopart-num}

>>>>

\begin{abstract}
The next generation of Gravitational wave detectors will accelerate the number
of ac{GW} detections such that we can gain new in site into the
physics behind the sources causing the phenomena. Numerical simulations and
matched filtering are the standard for detecting ac{GW}s for known
sources such as binary-black hole mergers. There are other sources of ac{GW}s
that remain elusive to standard modelling techniques and are expected to be
detectable. Here we construct a unmodelled burst generation scheme using
Generative adversarial networks - a powerful class of machine learning.
\chris{time to add a real abstract}
\end{abstract}

@@

\acrodef{GW}[GW]{Gravitational wave}
\acrodef{CBC}[CBC]{compact binary coalescence} 
\acrodef{ML}[ML]{machine learning}
\acrodef{GAN}[GAN]{generative adversarial network}
\acrodef{CGAN}[CGAN]{conditional generative adversarial network}
\acrodef{ACGAN}[ACGAN]{auxilliary conditional generative adversarial network}
\acrodef{DCGAN}[DCGAN]{deep convolutional generative adversarial network}
\acrodef{CNN}[CNN]{convolutional neural networks}
\acrodef{BBH}[BBH]{binary black hole}

@@

% introduce \ac{GW} astrophsyics
%
\ac{GW} astronomy is now an established field, starting with the first
detection of a binary black hole merger~\cite{Abbott2016} in September 2015.
Following this, the first and second observations runs (O1 and O2) of Advanced
LIGO and Advanced Virgo~\cite{Prospects-dets, AdvLIGO, AdvLIGO2, AdvVIRGO}
reported several more \ac{CBC} mergers~\cite{Abbott2016a, Abbott2017,
Abbott2017a, Abbott2017b}. On 17th August 2017 a binary neutron star merger was
observed alongside its electromagnetic counterpart for the first time, giving
rise to multi-messenger ac{GW} astronomy.

% introduce burst searches
%
With these successes and continued upgrades to the detectors, further
detections of \acp{CBC} are expected to be commonplace in future advanced
detcetor observation runs. Another group of \ac{GW} signals that has thus far
been undetected is \ac{GW} ``bursts". \ac{GW} bursts are classed as transient
signals of typically short duration ($<$ 1s) whose waveforms are not accurately
modelled or are complex to reproduce. Astrophysical sources for such transients
include: Core collapse supernova~\cite{Fryer_2003}, Pulsar
glitches~\cite{Andersson_2001}, Neutron star post-mergers~\cite{Baiotti_2007}
and other as-yet unexplained astrophysical phenomena.

% more details on burst searches
%
\ac{GW} detections so far have used a process called
matched-filtering,~\cite{Owen1998}, where a large template bank of possible
\ac{GW} waveforms are compared to the detector outputs. In order to increase
the chances of detection these template banks must span a large
multi-dimensional parameter space and the templates must be accurate, requiring
significant computational cost~\chris{the accuracy doesn't obviously lead to a
cost increase}. \ac{GW} bursts are un-modelled~\chris{except for the ones you
mentione before? SNe, glicthes, post-merger? }; therefore there are no
templates available and so matched-filtering is unsuitable for the detection of
these signals.  Instead, detection involves distinguishing the signal from
detector noise~\chris{but that's also what matched-filtering does}. Burst
searches look for excess power contained in the time-frequency
domain~\chris{all of them? In any case you have to provide references here to
the pipelines (cWB, x-pipeline, etc... Siong can help)} and rely on the
astrophysical burst waveform appearing in multiple detectors at similar times.
This is only possible if the detector noise is well characterised and the
candidate signal can be differentiated from systematic or environmental
glitches.

% Discuss the family of burst waveforms currently used and why - not in detail, just
% an introduction 
%
\ac{GW} burst detection algorithms~\cite{Klimenko_2008, Aso_2008} are tested
and tuned using modelled waveforms that may or may not have astrophysical
significance but have easy to define parameters and share characteristics of
real bursts that are enough to simulate a \ac{GW} passing between
detectors~\chris{long sentence}. Such waveforms include sine-Gaussians: a
Gaussian modulated sine wave that is characterised by its central frequency and
decay parameter. Bandlimited white-noise bursts: white noise that is contained
within a certain frequency range and ring-downs which mimic the damped
oscillations after a \ac{CBC} merger.~\chris{seems strange to list only 3 of
the 5 that we will be using later on.}

% cut this text if not going to be used
%
\begin{comment}Such waveforms may have
long-duration, short bandwidth (ringdowns), long-duration, large bandwidth
(inspirals) and many algorithms make use of sine-Gaussians: a Gaussian
modulated sine wave that is characterised by it's central frequency and narrow
bandwidth.~\chris{not the time to try to describe the types of burst waveforms.
Also, be careful with satying things like ringdowns have narrow bandwidth. I
know we specify a single frequency but becuase of the short duration, the
signal is broad band. Look at the FFT} This makes it a great tool for
diagnosing LIGOs sensitivity to frequency.~\chris{strange unfinished sentence.} 
\end{comment}

% Introduce ML techniques in GWs
%
With the expectation that there will be many more \ac{GW} detections in the
future, there is a growing need for fast and efficient \ac{GW} analysis methods
to match the rising number of detections. While still in its infancy, \ac{ML}
applications to \ac{GW} analyses have already shown great potential for
overcoming data bottlenecks~\chris{what does that mean?}. \ac{GW} analysis in
areas of detection~\cite{Gabbard2017,Gebhard_2019,Krastev_2020}, glitch
classification~\cite{Bahaadini, George_2018,Razzano_2018} and parameter
estimation~\cite{gabbard2019bayesian, shen2019deterministic,
green2020gravitationalwave}~\chris{unfinished sentence, plus I wouldn't
reference the Shen paper}. Applying these algorithms show similar results to
matched-filtering techniques and have the benefit of being orders of magnitude
faster. \michael{We mention aspects related to detection (pattern matching) but
also metion PE in the same sentence and I think this is a bit
confusing.}~\chris{yes, the matched-filtering comp[arison is only relevent for
searches. The biggest speed gain is for the Bayesian PE. I also think that you
need to reference more of the ML-GW literature here. I've added Mervyn's SNe
paper, Joe's latest paper, our latest Gaussian ?Mixture model paper, 2 CW
papers and the recent ML review paper to the bibligrapohy file but there are
quite a few others. Have a look at the review to steal references. In each case
try to say 1 sentence about what they do if possible.}

% What this paper does on GANs in 1 paragraph - the point of the paper
%
In this work we aim to explore the use of \ac{ML} to generate and interpret
unmodelled \ac{GW} burst waveforms. Using the generative machine learning
model: \acp{GAN}, we train on five classes of known~\chris{are they known? In
what sense?} burst morphologies~\chris{fancy word - what about waveform class?}
in the time domain. Working with the assumption that \acp{GAN} construct smooth
$n$-dimensional vector spaces between their input and output, we can then
explore the space between the five classes to construct new
unmodelled~\chris{hybrid?} waveforms. As all the computationally expensive
processes occur during training, the learned model after training is able to
produce waveforms in fractions of a second and in fact produce waveforms that
are impossible~\chris{never say never (or impossible)} to produce with current
techniques. These new varieties of waveforms can then be used to diagnose
detection algorithms and gain new insight into sources of \ac{GW}
bursts~\chris{yes, they can help diagnose or calibrate or assess the sensitivty
of detection algorithms but also allow us to better train our algorithms to a
broader range of possible signals and therefore enhance our detection ability}.

% the structure of the paper
%
This paper is organised as follows: \jordan{unstructured at the
moment}~\chris{time to add the structure.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this work we present the potential of Generative Adversarial Networks for burst ac{GW} analysis. We have shown that GANs have the ability to generate 5 class varieties of modelled burst \ac{GW} signals that can be generated at whim. The latent and class spaces were explored through interpolation and suggest that the space provides smooth translations between classes and overall waveform shape. We then showed targeted waveform generation by mixing classes to produce new unmodlled waveform varieties that can be used to test current burst search pipelines. \jordan{couple of sentences about classifier}.

@@

The approach shown in this work shows promise in generating unmodelled burst waveforms from exotic sources. Having the ability to quickly generate new waveforms is essential to test current detection schemes and their susceptibilty to unmodelled sources. We belive that GANs have the ability to generate high fidelity waveforms at a fraction of the computational expense and do not rely on large prior parameter space. Having banks of these waveforms at hand can aid in our understand of the physics processes behind these non-standard ac{GW} emitters.

@@

\section*{References}
\bibliography{references}
